---
title: "Education Data Analysis Project Proposal"
author: "cindyst2, please add your netID"
date: "July 11, 2018"
output: html_document
---

***
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```


## Dataset Description

- **Please add some descriptions to the data set, and other data sets if needed**
- https://www.cde.ca.gov/ta/ac/ap/apidatafiles.asp
- https://www.kaggle.com/passnyc/data-science-for-good#D5%20SHSAT%20Registrations%20and%20Testers.csv

## Proposal

- **Please add what questions you guys want to answer**
- How are student's performance different between charter schools vs none charter schools (cindyst2 we can probably compare with public vs privates schools if we can) 
- Does school funding, school district, teacher ratings play into any role?
- Does the background of the student such economic need index, student's race, family composition (single family cindyst2 no dataset for this yet) play a role with the student's performance? If so how much?
- Are there any unique programs in these schools that resulted in different student performances with similar backgrounds?

## Examine of Dataset

```{r}
library(readr)
#api_data = read_csv("14avgtx.txt") # cstseng: need to translate to csv

school = read_csv('2016 School Explorer.csv')
shsat = read_csv('D5 SHSAT Registrations and Testers.csv')

# Remove Adjusted Grade, New?, and Other Location Code
school = school[, c(-1, -2, -3)]
str(school, give.attr = FALSE)
school = subset(school, school$`Average Math Proficiency` !=  "N/A")
school$`Average Math Proficiency` = as.numeric(school$`Average Math Proficiency`)
school$`Average ELA Proficiency` = as.numeric(school$`Average ELA Proficiency`)

school = subset(school, school$`Economic Need Index` != "N/A")
school$`Economic Need Index` = as.numeric(school$`Economic Need Index`)

school$`Percent Asian` = as.numeric(school$`Percent Asian`)
school$`Percent Black` = as.numeric(school$`Percent Black`)
school$`Percent Black / Hispanic` = as.numeric(school$`Percent Black / Hispanic`)
school$`Percent Hispanic` = as.numeric(school$`Percent Hispanic`)
school$`Percent of Students Chronically Absent` = as.numeric(school$`Percent of Students Chronically Absent`)

math_model = lm(`Average Math Proficiency` ~ `Economic Need Index` , data = school)
ela_model = lm(`Average ELA Proficiency` ~ `Economic Need Index`, data = school)

summary(math_model)
plot(`Average Math Proficiency` ~ `Economic Need Index` , data = school, col = "blue")
plot(`Average ELA Proficiency` ~ `Economic Need Index` , data = school, col = "blue")
```

- todo: convert the 14avgtx.txt to excel, do some data clean up
- todo: clean up pyssnyc_data

***

##From Data-Sim guidelines

The names and NetIDs of the students who will be contributing to the group project
A tentative title for the project
Description of the dataset. You do not necessarily have to list all the variables, but at least mention those of greatest importance.
After reading the proposal, a reader should have a good understanding of your dataset without needing to reference the data itself or other documentation.
Background information on the dataset, including specific citation of its source
A brief statement of the business, science, research, or personal interest you have in the dataset that you hope to explore
Evidence that the data can be loaded into R. Load the data, and print the first few values of the response variable as evidence.
As a group, you will submit a .zip file as you would for homework that contains an .html and .Rmd file, as well as the data if it cannot be linked online. If your data is too large to submit and cannot be linked, please let us know and we will find an alternative.

